---
title: "Predicting 30-Day Readmission After Cardiac Surgery Using MIMIC-IV Dataset"
subtitle: "Biostat 203B and 212A Final Project"
author: "Fauzan Budi Prasetya and UID 306779006"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    self-contained: true
---

```{r}
# install.packages(c("bigrquery","DBI","tidyverse","glue"))
library(bigrquery)
library(DBI)
library(tidyverse)
library(glue)

# bigrquery::bq_auth(use_oob = TRUE)

con_bq <- DBI::dbConnect(
  bigrquery::bigquery(),
  project = "physionet-data-486705",
  dataset = "fauzan_staging",
  billing = "physionet-data-486705"
)

DBI::dbListTables(con_bq)
```

```{r}
## NEW VERSION
# my_staging <- "physionet-data-486705.fauzan_staging"
physio_hosp <- "physionet-data.mimiciv_3_1_hosp"
physio_ecg  <- "physionet-data.mimiciv_ecg"
physio_icu  <- "physionet-data.mimiciv_3_1_icu"

query <- "
    -- CREATE OR REPLACE TABLE `{my_staging}.cohort_table` AS
    
    -- STEP 1: Get patients's admissions with cardiac surgery (CSURG)
    WITH csurg_adm AS (
      SELECT
        a.subject_id, 
        a.hadm_id, 
        a.admittime, 
        a.dischtime,
        a.discharge_location, 
        a.admission_type,
        a.admission_location,
        a.insurance,
        a.language, 
        a.marital_status, 
        a.race,
        a.hospital_expire_flag,
        DATE_DIFF(dischtime, admittime, DAY) as hosp_los

      FROM `{physio_hosp}.admissions` a
      WHERE a.admittime IS NOT NULL
        AND a.dischtime IS NOT NULL
        AND EXISTS (
          SELECT 1
          FROM `{physio_hosp}.services` s
          WHERE s.hadm_id = a.hadm_id
            AND (s.curr_service = 'CSURG' OR s.prev_service = 'CSURG')
        )
    )

    -- STEP 2: Filter patient's admissions with ECG's check-up and take only first CSURG admission
    # , csurg_adm_with_ecg1 AS (
    , csurg_adm_with_ecg as (
      SELECT c.*
      FROM csurg_adm c
      WHERE EXISTS (
        SELECT 1
        FROM `{physio_ecg}.machine_measurements` e
        WHERE e.subject_id = c.subject_id
          AND e.ecg_time BETWEEN c.admittime AND c.dischtime
      )
    )

    , readmit_labels AS (

      -- STEP 3.1: Get all patient's admissions after previous CSURG and ECG
      WITH following_csurg_adm as (
        SELECT 
              a.subject_id, 
              a.admittime, 
              c.dischtime as prev_dischtime
          FROM `{physio_hosp}.admissions` a
          JOIN csurg_adm_with_ecg c 
            ON a.subject_id = c.subject_id 
            AND a.admittime > c.dischtime
        )

      -- STEP 3.2: Get the first following admission
      , first_following_csurg_adm as (
        SELECT
          subject_id,
          prev_dischtime,
          MIN(admittime) AS next_admittime
        FROM following_csurg_adm
        GROUP BY subject_id, prev_dischtime
      )
      
        -- STEP 3.2:  Label patient's readmission
        SELECT
          i.*,
          n.next_admittime,
          DATETIME_DIFF(n.next_admittime, i.dischtime, HOUR) AS hours_to_next,
          CASE
            WHEN n.next_admittime IS NULL THEN 0
            WHEN DATETIME_DIFF(n.next_admittime, i.dischtime, DAY) <= 30 
              AND DATETIME_DIFF(n.next_admittime, i.dischtime, HOUR) > 24 THEN 1
            ELSE 0
          END AS readmit_30d,
          CASE 
            WHEN n.next_admittime IS NOT NULL 
              AND DATETIME_DIFF(n.next_admittime, i.dischtime, HOUR) <= 24 THEN 'planned_or_transfer'
            ELSE 'standard'
          END as readmission_category
        FROM csurg_adm_with_ecg i
        LEFT JOIN first_following_csurg_adm n 
          ON n.subject_id = i.subject_id 
          AND n.prev_dischtime = i.dischtime
    )

    -- STEP 4: Exclude early readmissions, death and hospice and paliative care
    , final_cohort AS (
      SELECT * except(readmission_category)
      FROM readmit_labels
      WHERE 
        hospital_expire_flag = 0
        and discharge_location NOT IN ('HOSPICE', 'PALLIATIVE CARE')
        and readmission_category NOT IN ('planned_or_transfer')
    )

    -- STEP 5: Get patient's last ECG
    , last_ecg AS (
      # SELECT * EXCEPT(rn)
      SELECT * # edited because of this csurg_adm_with_ecg1
      FROM (
        SELECT
          c.hadm_id, 
          e.study_id, 
          e.ecg_time, 
          e.rr_interval, 
          # e.p_onset, 
          # e.p_end,
          e.qrs_onset, 
          e.qrs_end, 
          e.t_end, 
          # e.p_axis, 
          e.qrs_axis, 
          e.t_axis,
          SAFE_DIVIDE(60000, e.rr_interval) AS ecg_heart_rate,
          # e.p_end - e.p_onset AS p_dur,
          e.qrs_end - e.qrs_onset AS qrs_dur,
          e.t_end - e.qrs_onset AS qt_proxy,
          e.t_end - e.qrs_end AS t_dur,
          ROW_NUMBER() OVER (PARTITION BY c.hadm_id ORDER BY e.ecg_time DESC) AS rn
        FROM final_cohort c
        JOIN 
        (
          select 
              subject_id, 
              study_id, 
              ecg_time, 
              CASE WHEN rr_interval >= 10000 THEN NULL ELSE rr_interval END AS rr_interval,
              # CASE WHEN p_onset     >= 10000 THEN NULL ELSE p_onset     END AS p_onset,
              # CASE WHEN p_end       >= 10000 THEN NULL ELSE p_end       END AS p_end,
              CASE WHEN qrs_onset   >= 10000 THEN NULL ELSE qrs_onset   END AS qrs_onset,
              CASE WHEN qrs_end     >= 10000 THEN NULL ELSE qrs_end     END AS qrs_end,
              CASE WHEN t_end       >= 10000 THEN NULL ELSE t_end       END AS t_end,
              # CASE WHEN p_axis      >= 10000 THEN NULL ELSE p_axis      END AS p_axis,
              CASE WHEN qrs_axis    >= 10000 THEN NULL ELSE qrs_axis    END AS qrs_axis,
              CASE WHEN t_axis      >= 10000 THEN NULL ELSE t_axis      END AS t_axis
          from `{physio_ecg}.machine_measurements` 

        ) e
          ON e.subject_id = c.subject_id
          AND e.ecg_time BETWEEN c.admittime AND c.dischtime
      ) t
      WHERE rn = 1
    )

    -- STEP 6: Add icu stays of each patient's admission

    , icu_summary as (
      select 
        subject_id,
        hadm_id,
        sum(los) as los_icu
      FROM `{physio_icu}.icustays`
      group by 1,2
    )

    -- STEP 7: Add lab data of each patient's admission

    , lab_summary as (
    WITH base_labs AS (
      -- Filter for only the relevant IDs and normalize the naming
      SELECT 
        hadm_id, 
        charttime, 
        valuenum,
        CASE 
          -- High missingness (>40%) excluded:
          -- WHEN itemid IN (51002, 52642) THEN 'troponin_i'
          -- WHEN itemid = 51003 THEN 'troponin_t'
          -- WHEN itemid = 51921 THEN 'probnp_pleural'
          -- WHEN itemid = 50963 THEN 'ntprobnp'
          -- WHEN itemid = 50905 THEN 'ldl_calculated'
          -- WHEN itemid = 50906 THEN 'ldl_measured'
          -- WHEN itemid = 50903 THEN 'cholesterol_ratio'
          -- WHEN itemid = 50904 THEN 'hdl'
          -- WHEN itemid = 51000 THEN 'triglycerides'
          -- WHEN itemid = 52029 THEN 'ionized_calcium_percent'
          -- WHEN itemid = 50910 THEN 'ck_total'
          -- WHEN itemid IN (50911, 51595) THEN 'ck_mb'
          -- WHEN itemid = 51594 THEN 'ck_bb'
          -- WHEN itemid = 51596 THEN 'ck_mm'
          -- WHEN itemid = 50861 THEN 'alt'
          -- WHEN itemid = 50878 THEN 'ast'
          -- WHEN itemid IN (50862, 52022, 53085, 53138) THEN 'albumin'

          WHEN itemid IN (50971, 50822, 52452) THEN 'potassium'
          WHEN itemid IN (50983, 52455, 52623, 50824) THEN 'sodium'
          WHEN itemid IN (50893, 52034, 52035) THEN 'calcium_total'
          WHEN itemid IN (51624, 50808) THEN 'calcium_free'
          WHEN itemid = 50960 THEN 'magnesium'
          WHEN itemid IN (52024, 52546, 50912) THEN 'creatinine'
          WHEN itemid IN (52647, 51006) THEN 'bun'
          WHEN itemid = 51222 THEN 'hemoglobin'
          WHEN itemid IN (51221, 51638, 51639, 52028) THEN 'hematocrit'
          WHEN itemid IN (51301, 51755, 51756) THEN 'wbc'
          WHEN itemid IN (51265, 53189) THEN 'platelets'
          WHEN itemid = 50882 THEN 'bicarbonate'
          WHEN itemid IN (52500, 50868) THEN 'anion_gap'
          WHEN itemid IN (50813, 52442, 53154) THEN 'lactate'
          WHEN itemid IN (51237, 51675) THEN 'inrpt'
          WHEN itemid = 51274 THEN 'pt'
          WHEN itemid IN (50931, 52027, 52569, 50809) THEN 'glucose'
          WHEN itemid IN (50902, 52434, 52535, 50806) THEN 'chloride'
 
        END AS label
      FROM `{physio_hosp}.labevents`
      WHERE hadm_id IS NOT NULL
        AND valuenum IS NOT NULL
        AND itemid IN (
          50971, 50822, 52452, 50983, 52455, 52623, 50824, 50893, 51624, 
          52034, 50808, 52035, 50960, 52024, 52546, 50912, 52647, 51006, 
          51222, 51221, 51638, 51639, 52028, 51301, 51755, 51756, 51265, 
          53189, 50882, 52500, 50868, 50813, 52442, 53154, 51237, 51675, 
          51274, 50931, 52027, 52569, 50809, 50902, 52434, 52535, 50806
        )
    ),
    first_times AS (
      SELECT 
        hadm_id, 
        label, 
        MIN(charttime) AS first_charttime
      FROM base_labs
      GROUP BY hadm_id, label
    ),
    first_values AS (
      SELECT 
        bl.hadm_id, 
        bl.label, 
        AVG(bl.valuenum) AS first_val
      FROM base_labs bl
      INNER JOIN first_times ft
        ON bl.hadm_id = ft.hadm_id 
        AND bl.label = ft.label 
        AND bl.charttime = ft.first_charttime
      GROUP BY bl.hadm_id, bl.label
    )
    SELECT * FROM first_values
    PIVOT(
      AVG(first_val)
      FOR label IN (
        'potassium', 'sodium', 'calcium_total', 'calcium_free',
        'magnesium', 'creatinine', 'bun', 'hemoglobin', 'hematocrit', 
        'wbc', 'platelets', 'bicarbonate', 'anion_gap', 'lactate', 
        'inrpt', 'pt', 'glucose', 'chloride'
      )
    )
    )

    -- STEP 8: Add ECG features of each patients
    SELECT
      # c.* except(rn),
      c.*,
      l.study_id,
      l.ecg_time,
      l.rr_interval,
      # l.p_onset,
      # l.p_end,
      l.qrs_onset,
      l.qrs_end,
      l.t_end,
      # l.p_axis,
      l.qrs_axis,
      l.t_axis,
      l.ecg_heart_rate,
      # p_dur,
      qrs_dur,
      qt_proxy,
      t_dur,
      p.gender,
      EXTRACT(YEAR FROM admittime) - p.anchor_year + p.anchor_age as age,
      case when los_icu is null then 0 else los_icu end as los_icu,
      lb.* except(hadm_id)

    FROM final_cohort c
    LEFT JOIN last_ecg l
      ON l.hadm_id = c.hadm_id
    LEFT JOIN `{physio_hosp}.patients` p 
      ON c.subject_id = p.subject_id
    LEFT JOIN icu_summary i
      ON i.hadm_id = c.hadm_id
    LEFT JOIN lab_summary lb
      ON lb.hadm_id = c.hadm_id
      ;
"

cohort <- dbGetQuery(con_bq, glue(query))

```

```{r}
library(comorbidity)

query_diag <- "
  SELECT
    d.subject_id,
    d.hadm_id,
    d.icd_code,
    d.icd_version
  FROM `{physio_hosp}.diagnoses_icd` d
  WHERE d.icd_code IS NOT NULL;
  "
dx <- dbGetQuery(con_bq, glue(query_diag)) |>
  filter(hadm_id %in% cohort$hadm_id)

dx9  <- dx |> filter(icd_version == 9)
dx10 <- dx |> filter(icd_version == 10)

elix9 <- if (nrow(dx9) > 0) {
  comorbidity(
    x = dx9,
    id = "hadm_id",
    code = "icd_code",
    map = "elixhauser_icd9_quan",
    assign0 = TRUE,
    tidy.codes = TRUE
  )
} else {
  NULL
}

elix10 <- if (nrow(dx10) > 0) {
  comorbidity(
    x = dx10,
    id = "hadm_id",
    code = "icd_code",
    map = "elixhauser_icd10_quan",
    assign0 = TRUE,
    tidy.codes = TRUE
  )
} else {
  NULL
}

elix_all <- bind_rows(elix9, elix10) |>
  group_by(hadm_id) |>
  summarise(across(where(is.numeric), ~max(.x, na.rm = TRUE)), .groups = "drop")

nrow(elix_all)
n_distinct(elix_all$hadm_id)
```

```{r}

cohort_final <- cohort |>
  left_join(elix_all, by = "hadm_id") |>
  mutate(
    admission_type = fct_lump_n(admission_type, n = 5, other_level = "Other"),
    admission_location = fct_lump_n(admission_location, n = 5, other_level = "Other"),
    discharge_location = fct_lump_n(discharge_location, n = 5, other_level = "Other"),
    language = fct_lump_n(language, n = 5, other_level = "Other"),
    race = fct_collapse(
      race,
      ASIAN = c(
        "ASIAN", 
        "ASIAN - ASIAN INDIAN", 
        "ASIAN - CHINESE", 
        "ASIAN - KOREAN", 
        "ASIAN - SOUTH EAST ASIAN"
      ),
      BLACK = c(
        "BLACK/AFRICAN", 
        "BLACK/AFRICAN AMERICAN", 
        "BLACK/CAPE VERDEAN", 
        "BLACK/CARIBBEAN ISLAND"
      ),
      HISPANIC = c(
        "HISPANIC OR LATINO", 
        "HISPANIC/LATINO - CENTRAL AMERICAN", 
        "HISPANIC/LATINO - COLUMBIAN", 
        "HISPANIC/LATINO - CUBAN", 
        "HISPANIC/LATINO - DOMINICAN", 
        "HISPANIC/LATINO - GUATEMALAN", 
        "HISPANIC/LATINO - HONDURAN", 
        "HISPANIC/LATINO - MEXICAN", 
        "HISPANIC/LATINO - PUERTO RICAN", 
        "HISPANIC/LATINO - SALVADORAN",
        "SOUTH AMERICAN"
      ),
      WHITE = c(
        "WHITE", 
        "WHITE - BRAZILIAN", 
        "WHITE - EASTERN EUROPEAN", 
        "WHITE - OTHER EUROPEAN", 
        "WHITE - RUSSIAN"
      ),
      Other = c(
        "AMERICAN INDIAN/ALASKA NATIVE", 
        "MULTIPLE RACE/ETHNICITY", 
        "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER", 
        "OTHER", 
        "PATIENT DECLINED TO ANSWER", 
        "PORTUGUESE", 
        "UNABLE TO OBTAIN", 
        "UNKNOWN"
      ))
    )

nrow(cohort_final)
dplyr::n_distinct(cohort_final$subject_id)
```

```{r}
write_csv(cohort_final, "cohort_final.csv")
```

```{r}
cohort_final <- cohort_final |>
  mutate(readmit_30d = factor(readmit_30d, levels = c("0", "1")))

str(cohort_final)
summary(cohort_final)
```

```{r}
str(cohort_final)
nrow(cohort_final)
ncol(cohort_final)
n_distinct(cohort_final$hadm_id)
n_distinct(cohort_final$subject_id)
```

```{r}
library(gtsummary)
desc_table2 <- cohort_final |> 
  select(-subject_id, -hadm_id, -study_id, 
    -next_admittime, -admittime, -dischtime, 
    -ecg_time, -hours_to_next, -hospital_expire_flag) |> 
  tbl_summary(by = readmit_30d, missing = "ifany") |> 
  add_overall() |> 
  add_n() |> 
  bold_labels()

desc_table2
```

```{r}
split <- initial_split(cohort_final, prop = 0.8, strata = readmit_30d)
train_data <- training(split)
test_data  <- testing(split)
```


```{r}
library(recipes)
set.seed(123)

rec <- recipe(readmit_30d ~ ., data = train_data) |>
  
  # remove IDs and datetime variable
  step_rm(subject_id, hadm_id, study_id, next_admittime, admittime, dischtime, ecg_time, hours_to_next, hospital_expire_flag) |>

  # handle categorical variables
  step_string2factor(all_nominal_predictors()) |>
  step_other(all_nominal_predictors(), threshold = 0.05) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  
  # remove constant and near zero variance predictors
  step_zv(all_predictors()) |>
  step_nzv(all_predictors()) |>  
  
  # handle numerical variables
  # step_corr(all_numeric_predictors(), threshold = 0.9) |>
  step_impute_knn(all_numeric_predictors()) |> #impute missing values for numeric variables using KNN
  step_normalize(all_numeric_predictors())
  # |>
  # step_smote(readmit_30d) # handle class imbalance
```

```{r}
# prepped_data <- prep(rec)
# cohort_final_cleaned <- bake(prepped_data, new_data = NULL)
# summary(cohort_final_cleaned)
# str(cohort_final_cleaned)
```

```{r}
xgb_model <- boost_tree(
  learn_rate = tune(),
  tree_depth = tune(),
  trees = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

rf_model <- rand_forest(
  mtry = tune(),    
  trees = tune(),    
  min_n = tune()     
) %>%
  set_engine("randomForest") %>% 
  set_mode("classification")

log_model <- logistic_reg(
  penalty = tune(),   
  mixture = tune()    
) %>%
  set_engine("glmnet") %>%
  set_mode("classification")


```

```{r}
wf_log <- workflow() %>%
  add_recipe(rec) %>%
  add_model(log_model)
  
wf_rf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_model)

wf_xgb <- workflow() %>%
  add_recipe(rec) %>%
  add_model(xgb_model)
```


```{r}
folds <- vfold_cv(train_data, v = 5, strata = readmit_30d)
```

```{r}

log_grid <- grid_regular(
  penalty(range = c(-4, 0)),   
  mixture(range = c(0, 1)),    
  levels = 5
)

rf_grid <- grid_regular(
  mtry(range = c(1L, 20L)),    
  trees(range = c(50L, 500L)),
  min_n(range = c(1L, 20L)),
  levels = 5
)

xgb_grid <- grid_regular(
  learn_rate(range = c(-2, -1)),
  tree_depth(range = c(1L, 10L)),
  trees(range = c(50L, 500L)),
  levels = 5
)

```

```{r}
library(future)
plan(multisession, workers = parallel::detectCores() - 1)

ctrl <- control_grid(
  verbose = TRUE,
  allow_par = TRUE,
  parallel_over = "everything"  
)

tune_res_log <- tune_grid(
  wf_log,
  resamples = folds,
  grid = log_grid,
  metrics = metric_set(roc_auc),
  control = ctrl
)

tune_res_rf <- tune_grid(
  wf_rf,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(roc_auc),
  control = ctrl
)

tune_res_xgb <- tune_grid(
  wf_xgb,
  resamples = folds,
  grid = xgb_grid,
  metrics = metric_set(roc_auc),
  control = ctrl
)
```

```{r}
best_params_log <- select_best(tune_res_log, metric = "roc_auc")

all_results_log <- tune_res_log %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  select(penalty, mixture, mean, .metric, .estimator, n, std_err, .config) %>%
  arrange(desc(mean))

best_params_rf <- select_best(tune_res_rf, metric = "roc_auc")

all_results_rf <- tune_res_rf %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  select(mtry, trees, min_n, mean, .metric, .estimator, n, std_err, .config) %>%
  arrange(desc(mean))

best_params_xgb <- select_best(tune_res_xgb, metric = "roc_auc")

all_results_xgb <- tune_res_xgb %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  select(learn_rate, tree_depth, trees, mean, .metric, .estimator, n, std_err, .config) %>%
  arrange(desc(mean))

print("Logistic Regression Results:")
print(all_results_log, n = nrow(all_results_log))
print("Random Forest Results:")
print(all_results_rf, n = nrow(all_results_rf))
print("XGBoost Results:")
print(all_results_xgb, n = nrow(all_results_xgb))

```

```{r}
# Logistic Regression Line plot: ROC-AUC vs Penalty for different mixture values
plot_log <- ggplot(all_results_log, aes(x = penalty, y = mean, color = factor(mixture), 
                              group = mixture)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(title = "Logistic Regression ROC-AUC vs Penalty Across Mixture Values",
       x = "Penalty (log scale)", y = "ROC-AUC",
       color = "Mixture") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_viridis_d()

print(plot_log)

# Random Forest Boxplot: Distribution of ROC-AUC by trees
plot_rf <- ggplot(all_results_rf, aes(x = factor(trees), y = mean, fill = factor(trees))) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
  scale_fill_viridis_d() +
  labs(title = "Random Forest ROC-AUC Distribution by Number of Trees",
       x = "Number of Trees", y = "ROC-AUC") +
  theme_minimal() +
  theme(legend.position = "none")

print(plot_rf)

# XGBoost Line plot: ROC-AUC vs Learn Rate for different Tree Depths
plot_xgb <- ggplot(all_results_xgb, aes(x = learn_rate, y = mean, color = factor(tree_depth), group = tree_depth)) +
  geom_smooth(se = FALSE, method = "loess", size = 1) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_x_log10() +
  scale_color_viridis_d(name = "Tree Depth") +
  labs(title = "XGBoost ROC-AUC vs Learn Rate Across Different Tree Depths",
       x = "Learn Rate (log scale)", y = "ROC-AUC") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(plot_xgb)
```

```{r}
write_csv(all_results_log, file = "all_results_log.csv")
write_csv(all_results_rf, file = "all_results_rf.csv")
write_csv(all_results_xgb, file = "all_results_xgb.csv")
```


```{r}
my_metrics <- metric_set(
  roc_auc,
  pr_auc,
  accuracy,
  ppv,
  npv
)

final_wf_log <- finalize_workflow(wf_log, best_params_log)
final_wf_rf <- finalize_workflow(wf_rf, best_params_rf)
final_wf_xgb <- finalize_workflow(wf_xgb, best_params_xgb)

final_fit_log <- final_wf_log %>%
  last_fit(split, metrics = my_metrics)
final_fit_rf <- final_wf_rf %>%
  last_fit(split, metrics = my_metrics)
final_fit_xgb <- final_wf_xgb %>%
  last_fit(split, metrics = my_metrics)

print("Final Logistic Regression Metrics:")
collect_metrics(final_fit_log)
print("Final Random Forest Metrics:")
collect_metrics(final_fit_rf)
print("Final XGBoost Metrics:")
collect_metrics(final_fit_xgb)
```

```{r}
library(yardstick)

# Collect predictions from all models
predictions_log <- final_fit_log %>%
  collect_predictions() %>%
  mutate(model = "Logistic Regression")

predictions_rf <- final_fit_rf %>%
  collect_predictions() %>%
  mutate(model = "Random Forest")

predictions_xgb <- final_fit_xgb %>%
  collect_predictions() %>%
  mutate(model = "XGBoost")

# Combine all predictions
all_predictions <- bind_rows(predictions_log, predictions_rf, predictions_xgb)

# Calculate AUC for each model
auc_values <- all_predictions %>%
  group_by(model) %>%
  roc_auc(truth = readmit_30d, .pred_0) %>%
  mutate(auc_label = paste(model, "- AUC =", round(.estimate, 3)))

# Create ROC curves for all models
roc_data <- all_predictions %>%
  group_by(model) %>%
  roc_curve(truth = readmit_30d, .pred_0)

# Plot combined ROC curves
auc_curve_plot <- ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(size = 1.2) +
  geom_abline(linetype = "dashed", color = "gray50", alpha = 0.8) +
  coord_equal() +
  scale_color_manual(
    values = c("Logistic Regression" = "#2E86AB", 
               "Random Forest" = "#A23B72", 
               "XGBoost" = "#F18F01"),
    labels = c(
      paste0("Logistic Regression (AUC = ", 
             round(filter(auc_values, model == "Logistic Regression")$.estimate, 3), ")"),
      paste0("Random Forest (AUC = ", 
             round(filter(auc_values, model == "Random Forest")$.estimate, 3), ")"),
      paste0("XGBoost (AUC = ", 
             round(filter(auc_values, model == "XGBoost")$.estimate, 3), ")")
    )
  ) +
  labs(
    title = "ROC Curves - Model Comparison",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    color = "Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

``` {r}
library(ggplot2)
library(dplyr)
library(broom)

final_model_log <- extract_fit_parsnip(final_fit_log$.workflow[[1]])

importance_data <- tidy(final_model_log) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    odds_ratio = exp(estimate),
    direction = ifelse(odds_ratio > 1, "Increase Odds", "Decrease Odds"),
    label_text = round(odds_ratio, 2)
  ) %>%
  group_by(direction) %>%
  slice_max(abs(estimate), n = 10) %>%
  ungroup()

ggplot(importance_data, aes(x = reorder(term, odds_ratio), y = odds_ratio, fill = direction)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = label_text), 
            hjust = ifelse(importance_data$odds_ratio > 1, -0.1, 1.1), 
            size = 3.5, 
            fontface = "bold") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  coord_flip() +
  scale_y_log10(expand = expansion(mult = c(0.2, 0.2))) + 
  scale_fill_manual(values = c("Increase Odds" = "#e67e22", "Decrease Odds" = "#3498db")) +
  labs(
    title = "Feature Importance: Odds Ratios",
    subtitle = "Blue: Higher odds | Orange: Lower odds (Reference line at 1.0)",
    x = NULL,
    y = "Odds Ratio (Log Scale)"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold")
  )
```

```{r}
library(vip)

final_model_rf <- extract_fit_parsnip(final_fit_rf$.workflow[[1]])
final_model_xgb <- extract_fit_parsnip(final_fit_xgb$.workflow[[1]])

vip(final_model_rf)
vip(final_model_xgb)
```


# 1. Title Page
- Project title: "Predicting 30-Day Readmission After Cardiac Surgery Using MIMIC-IV Dataset"
- Your team members: Yash Motwani, Sofia Landinez, and Mert Cumhur

# 2. Abstract
Provide a succinct summary of your entire project (~250–300 words), including:
- Background – Context and why the problem matters
- Objectives – What your project intends to investigate
- Methods – Main approach and data sources
- Results – Key findings
- Conclusion – Implications and potential impact

Abstract
Background:  Hospital readmission within 30 days of discharge is a common among patients undergoing cardiac surgery. By indentifying patients at high risk for readmission, healthcare providers can implement targeted interventions to improve post-discharge care and reduce preventable readmissions.

Objectives: The goal of this project is to develop and validate predictive models for 30-day hospital readmission following cardiac-surgery patients who underwent an EKG during their hospital stay. 

Methods: 
This study was conducted utilizing MIMIC-IV datasets that are focused on electronic health record data from cardiac-surgery patients with an EKG examination. This include patient demographics, hospital stay details, comorbidities, EKG features, and laboratory results. 7000 patients of cardiac surgery were included in the final cohort, with 15% experiencing readmission within 30 days. Three predictive classification models were developed: a logistic regression, a random forest, and an XGBoost model. Performance was evaluated using the area under the receiver operating characteristic curve (ROC AUC), with feature importance assessed to identify key predictors.

Results: The XGBoost model achieved the highest performance, resulting a ROC AUC of 0.632. This moderate performance suggests that while readmission risk among cardiac surgery patient is complex, structured medical record data from EKGs and the associated hospital stay contain a measurable signal. Preliminary feature importance analysis indicated that elements length of stay, specific EKG abnormalities, and demographics were among the strongest predictors.

Conclusion: This study demonstrates that machine learning models using readily available clinical data can moderately predict 30-day readmission in high-risk cardiac-surgery patients with an EKG. A ROC AUC of 0.632 is starting point for identifying patients who may benefit from enhanced discharge planning. Future work to improve predictive performance will focus on integrating hospital recharge note, EKG results note from physicians, and a more comprehensive set of social determinants of health.





# 3. Introduction
Establish your project’s rationale and research gap.

## 3.1 Background

  Hospital readmission within 30 days of discharge is happening in approximately 15-20% of patients undergoing cardiac surgery in the US. It is associated with increased healthcare costs, patient morbidity, and mortality.  The Centers for Medicare and Medicaid Services Readmission Reduction Program was initiated on October 1, 2012, as part of the Affordable Care Act. This program administers penalties of Medicare revenue for an excess 30-day readmission rate in the categories of heart failure and myocardial infarction.
  
  Identifying patients at high risk for readmission can enable healthcare providers to implement targeted interventions, such as enhanced discharge planning, post-discharge follow-up, and patient education, to improve outcomes and reduce preventable readmissions. Despite the importance of this issue, accurately predicting which patients are at risk for readmission remains a challenge. Previous studies have explored various clinical and demographic factors associated with readmission risk, but there is a need for more comprehensive models that integrate structured data from electronic health records (EHRs), including EKG features and hospital stay details, to improve predictive performance. This project aims to fill this gap by developing and validating machine learning models that utilize structured EHR data to predict 30-day readmission after cardiac surgery.

  The hypothesis of this project is that machine learning models can predict 30-day readmission after cardiac surgery using structured data from EKGs and associated hospital stay details, with the expectation that the models will achieve a ROC AUC of at least 0.65, indicating the typical performance of such readmission prediction models.

  https://www.cms.gov/medicare/payment/prospective-payment-systems/acute-inpatient-pps/hospital-readmissions-reduction-program-hrrp 
  https://pmc.ncbi.nlm.nih.gov/articles/PMC7382395
  https://pmc.ncbi.nlm.nih.gov/articles/PMC4186890/ 

# 4. Methods


## 4.1 Data Source or Materials

This study was conducted utilizing MIMIC-IV datasets that are focused on electronic health record data from cardiac-surgery patients with an inpatient EKG. The datasets were accessed through BigQuery, and the analysis was performed using R. The study cohort included adult patients who underwent cardiac surgery and had at least one EKG examination during their hospital stay. The exclusion criteria included patients who died during the index hospitalization, those with missing discharge timestamps, and those who were discharged to hospice care or palliative care. There are 8167 admissions of 7640 patients of cardiac surgery included in the final cohort, with 15% experiencing readmission within 30 days. The data access and analysis were conducted in compliance with the MIMIC-IV data use agreement, ensuring patient privacy and appropriate use of the data. All analyses were performed on de-identified data, and the study was approved by the relevant institutional review board, and can be accessed through BigQuery after submitting a data use request.

## 4.2 Variables / Features

There are total 70 features included in the final model, which are patient demographics, hospital stay details, comorbidities, EKG features, and laboratory results. Demographics data are taken from the patient records. They are consumed to understand the socioeconomic and health background of the patients, which can influence readmission risk. Hospital stay details are taken from the admissions table, including length of stay, admission type, admission location, discharge location, insurance. To get length of stay in ICU, we consume ICU stays table. Hospital stay details are important because they can reflect the severity of illness and the complexity of care, which are often associated with readmission risk. Comorbidity features were derived using the Elixhauser comorbidity index, which is a widely used method for quantifying patient comorbidities based on ICD diagnosis codes. Comorbidities are important predictors of readmission risk, as patients with multiple or severe comorbid conditions are more likely to experience complications and require readmission. EKG features were extracted from machine measurements table, and take the last EKG performed during the hospital stay, between admitted and discharged. After excluding raw EKG column that have high rate of missing data, derived features are calculated, such as heart rate, QRS duration, QT interval, and T wave abnormalities. These features are relevant because they can indicate underlying cardiac conditions and abnormalities that may contribute to readmission risk. Laboratory results included key biomarkers that related with cardiac issues, such as potassium, sodium, calcium, creatinine, and hemoglobin levels. They are important predictors of readmission risk, as abnormal values can indicate ongoing health issues or complications that may lead to readmission.

The features were processed using a combination of imputation for missing values, normalization for numerical variables, and encoding for categorical variables. For example, categorical variables were transformed using one-hot encoding, while numerical variables were normalized to ensure they were on a comparable scale for modeling. The processing steps were implemented using the `recipes` package in R, which allowed for a systematic and reproducible approach to data preprocessing.

## 4.3 Analytical Approach / Modeling

The index time was defined as the discharge time of the cardiac surgery admission, with a prediction window of 30 days for readmission. The outcome variable was defined as a binary indicator of whether the patient was readmitted to the hospital within 30 days of discharge from the index cardiac surgery admission. They were discharge under 24 hours were not considered as readmission.

Three predictive classification models were developed: a logistic regression model, a random forest, and an XGBoost. Model were developed in R, utilizing package mainly tidyverse and tidymodels workflow. After that, performance was evaluated using the area under the receiver operating characteristic curve (ROC AUC), with feature importance assessed to identify key predictors.

All the parameters tuning are using grid search with 5-fold cross validation. The best model is selected based on the highest ROC AUC. In logistic regression, we used regularization strength (penalty or lambda) to control the amount of flexibility applied to prevent overfitting and the elastic net mixing parameter (mixture or alpha) which is a number between 0 and 1, where 0 corresponds to pure ridge regression and 1 corresponds to pure lasso regression, to balance between two types of regularization and achieve better predictive performance. 

For random forest, we tuned the feature randomness (mtry), tree complexity (min_n), and ensemble size which is number of trees (trees). These three parameters are chosen because they are the core parameters, as stated in the official documentation of randomForest package and other research paper by Ishwaran, that control the randomness, complexity, and size of the random forest model, which are crucial for optimizing its performance and preventing overfitting.

For XGBoost, we tuned the learning rate, maximum depth, and number of trees. They are the core parameters that control the learning process, model complexity, and ensemble size of the XGBoost model, which are essential for optimizing its performance and preventing overfitting. They are chosen based on the official documentation of xgboost package and other research paper by Chen and Guestrin.

https://xgboost.readthedocs.io/en/latest/tutorials/param_tuning.html?from_column=20423&from=20423#control-overfitting 
https://parsnip.tidymodels.org/reference/rand_forest.html
https://zbmath.org/?q=an:07260271 
https://iwaponline.com/view-large/3703360 
https://arxiv.org/abs/1603.02754 

## 4.4 Model Evaluation
- Performance metrics (e.g., ROC/AUC, accuracy)
- Validation methods (train/test split, cross-validation, etc.)

After training those three models with 5-fold cross-validation and utilizing hyperparameter tuning, we evaluated their performance on the test set. The primary performance metric used for model evaluation was the area under the receiver operating characteristic curve (ROC AUC), which measures the model's ability to discriminate between patients who were readmitted and those who were not. Additionally, we calculated other metrics such as accuracy, ppv, and npv to provide a comprehensive assessment of model performance. The final models were compared based on their ROC AUC values, with the best-performing model selected for further analysis and interpretation. 

# 5. Results
State your findings clearly and objectively. Provide tables and figures where appropriate.

## 5.1 Cohort / Sample Characteristics
- Description of your data (counts, demographics, etc.)

```{r}
desc_table2
```

## 5.2 Model Performance
- Summary of main results (metrics, comparisons among models)

XGBoost achieved the highest ROC AUC of 0.643, indicating moderate predictive performance in identifying patients at risk for 30-day readmission after cardiac surgery, followed by logistic regression with a ROC AUC of 0.642, and random forest with a ROC AUC of 0.634. The ROC curves for all three models are shown in the figure below, with the AUC values indicated in the legend.

```{r}
auc_curve_plot
```

## 5.3 Additional Analyses
- Feature importance (Shapley values), calibration, etc.


# 6. Discussion
Interpret the results and place them in context.

## 6.1 Summary of Findings
- What the results mean and how they relate to your aims

## 6.2 Comparison with Previous Work
- How your findings align with or differ from others

## 6.3 Strengths and Limitations
- What worked well; what constraints or biases exist

## 6.4 Implications
- Practical, theoretical, or clinical applications of your findings

# 7. Conclusion
Wrap up with clear takeaways and next steps.
- Recap main conclusions
- Significance and contributions
- Suggestions for future research or improvements

# 8. References
- List all sources cited in your paper
- Use consistent citation formatting (e.g., APA, MLA, or journal style)

# 9. AI usage acknowledgement
This project utilized AI tools (ChatGPT, Gemini, and Deepseek) in github copilot and web browser for code generation, debugging, writing assistance. The AI was used to brainstorm challenges and generate initial drafts. All outputs from the AI were reviewed and validated by the researcher to ensure accuracy and relevance to the project objectives.